\LARGE \bf Mean Estimation from One-Bit Measurements

### 2021-09-29 uniform-are-weirdos.tex
fixed all comments after first revision, sending to John

Here we consider the uniform distribution as our location family,
demonstrating that in the adaptive setting~\eqref{item:adaptive} or even the
one-step adaptive setting~(\ref{item:one-step-adaptive}'), constrained
estimators can attain rates faster than the $1 / \sqrt{n}$ rates regular
estimands allow. Indeed, define $c(x) = -\log 2$ for $x \in [-1, 1]$ and
$c(x) = -\infty$ for $x \not \in [-1, 1]$. Then $f(x) = e^{-c(x)}$ is
log-concave and symmetric, and we may consider the location family with
densities $f(x - \theta)$. For notational simplicity, we assume we have
a sample of size $2n$. We provide a proof sketch that
there is a one-step adaptive estimator $\theta_n$ such that
\begin{equation}
  \label{eqn:uniforms-are-easy}
  \sup_{|\theta| \le \log n}
  P_\theta\left(|\theta_n - \theta| \ge \frac{8 \log n}{n^{3/4}}\right)
  \le \frac{2}{n^2}.
\end{equation}
for all large $n$,
and so (by the Borel-Cantelli lemmas), for any $\theta \in \R$ we have
$P_\theta(|\theta_n - \theta| \le \sqrt{2 \log n} / n^{3/4} ~
\mbox{eventually}) = 1$. This is of course faster than the $1/\sqrt{n}$
rates we prove throughout.

To prove inequality~\eqref{eqn:uniforms-are-easy}, we proceed in two steps,
both quite similar.
First, we define an initial estimator $\theta\init_n$.
Let $\epsilon > 0$, which we will determine presently, though we will
take $n \epsilon \to \infty$ as $n \to \infty$, so that we may assume
w.l.o.g.\ that $\theta \in [-n\epsilon/2, n \epsilon/2]$. Take the interval
$[-n\epsilon, n\epsilon]$, and construct
$m$ thresholds at intervals of size $2 n \epsilon / m$; let
the $j$th such threshold be
\begin{equation*}
  t_j \defeq -n \epsilon + \frac{2 n (j-1) \epsilon}{m}
\end{equation*}
Then we ``assign'' observations to each pair of thresholds, so that
threshold $j$ corresponds to observations $I_j \defeq \{\frac{n (j - 1)}{m}
+ 1, \ldots, \frac{n j}{m}\}$, of which there are $n/m$.  For each index $i
\in I_j$, we set
\begin{equation*}
  B_i = \begin{cases}
    1 & \mbox{if~} X_i \ge t_j \\
    0 & \mbox{otherwise}.
  \end{cases}
\end{equation*}
Then we simply set $\theta_n\init$ to be the maximal threshold for which
$B_i = 0$ for all observations $X_i$ corresponding to that threshold.

Let us now consider the probability that $\theta_n\init$ is substantially
wrong. For notational simplicity, let $U_i = (1 + X_i - \theta) / 2$, so that
the $U_i$ are uniform on $[0, 1]$.
First, note that we always have $\theta_n\init \ge \theta
- \frac{2 n \epsilon}{m}$, because no observations will be below
the appropriate threshold. Let $j\opt$ be the smallest index $j$
for which $\theta \le t_{j\opt}$, and consider
the index sets $I_{j\opt}$, $I_{j\opt + 1}$, and so on.
The event $\theta\init_n \ge t_{j\opt} + \frac{2 n \epsilon}{m}$ may
occur only if for each of the $n/m$ observations in the set
$I_{j\opt + 1}$, we have $U_i \ge \frac{n \epsilon}{m}$. Thus,
\begin{equation*}
  P_\theta\left(\theta_n\init \ge t_{j\opt} + \frac{2 n \epsilon}{m}
  \right)
  \le \left(1 - \frac{n \epsilon}{m}\right)^\frac{n}{m}
  \le \exp\left(-\frac{n^2\epsilon}{m^2}\right).
\end{equation*}
Setting the number of bins $m = \sqrt{n}$,
the resolution $\epsilon = 2 \log n / n$, we obtain
$P_\theta(\theta_n\init \ge t_{j\opt} + 4 \log n / \sqrt{n})
\le n^{-2}$. Thus we have
\begin{equation}
  \label{eqn:quality-of-initial-estimate}
  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^2}.
\end{equation}

The second stage estimator follows roughly the same strategy, except that
the resolution of the bins is tighter. In particular, let us assume that
$|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$, which happens
eventually by inequality~\eqref{eqn:quality-of-initial-estimate}.  (We will
assume this tacitly for the remainder of the argument.)  Consider the
interval $\Theta_n \defeq \theta_n\init + [-\frac{16 \log n}{\sqrt{n}},
  \frac{16 \log n}{\sqrt{n}}]$ centered at $\theta_n\init$; we know that the
interval includes $[\theta - \frac{8 \log n}{\sqrt{n}}, \theta + \frac{8
    \log n}{\sqrt{n}}]$. Without loss of generality we assume $\theta_n\init
= 0$.  Following precisely the same discretization strategy as that for
$\theta_n\init$, we divide $\Theta_n$ into $m$ equal intervals, with
thresholds $t_j = -\frac{16 \log n}{\sqrt{n}} + \frac{32 (j - 1) \log n}{m
  \sqrt{n}}$; let $\epsilon_n = \frac{32 \log n}{m \sqrt{n}}$ be the width of
these intervals.  Then following exactly the same reasoning as above, we
assign indices $I_j = \{\frac{n(j - 1)}{m} + 1, \ldots, \frac{n j}{m}\}$ and
for $i \in I_j$, set $B_i = 1$ if $X_i \ge t_j$. We define $\theta_n$ to be
the maximal threshold $t_j$ for which $B_i = 0$ for all observations $X_i
\in I_j$. Then following precisely the reasoning above, we have (on the
event that $|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$)
\begin{equation*}
  P_\theta(|\theta_n - \theta|
  \ge 2 \epsilon_n)
  \le (1 - \epsilon_n)^\frac{n}{m}
  \le \exp\left(-\frac{n \epsilon_n}{m}\right)
  = \exp\left(-\frac{32 \sqrt{n} \log n}{m^2}\right).
\end{equation*}
Set $m = 4 n^{1/4}$ to obtain the claimed
result~\eqref{eqn:uniforms-are-easy}.

>>>>

Here we consider the uniform distribution as our location family,
demonstrating that in the adaptive setting~\eqref{item:adaptive} or even the
one-step adaptive setting~(\ref{item:one-step-adaptive}'), constrained
estimators can attain rates faster than the $1 / \sqrt{n}$ rates regular
estimands allow. Indeed, define $c(x) = -\log 2$ for $x \in [-1, 1]$ and
$c(x) = -\infty$ for $x \not \in [-1, 1]$. Then $f(x) = e^{-c(x)}$ is
log-concave and symmetric, and we may consider the location family with
densities $f(x - \theta)$. For notational simplicity, we assume we have
a sample of size $2n$. We provide a proof sketch that
there is a one-step adaptive estimator $\theta_n$ such that
\begin{equation}
  \label{eqn:uniforms-are-easy}
  \sup_{|\theta| \le \log n}
  P_\theta\left(|\theta_n - \theta| \ge \frac{\newtext{16} \log n}{n^{3/4}}\right)
  \le \frac{2}{n^2}.
\end{equation}
for all large $n$,
and so (by the Borel-Cantelli lemmas), for any $\theta \in \R$ we have
$P_\theta(|\theta_n - \theta| \le \newtext{16 \log n} / n^{3/4} ~
\mbox{eventually}) = 1$. This is of course faster than the $1/\sqrt{n}$
rates we prove throughout.

To prove inequality~\eqref{eqn:uniforms-are-easy}, we proceed in two steps,
both quite similar.
First, we define an initial estimator $\theta\init_n$.
Let $\epsilon > 0$, which we will determine presently, though we will
take $n \epsilon \to \infty$ as $n \to \infty$, so that we may assume
w.l.o.g.\ that $\theta \in [-n\epsilon/2, n \epsilon/2]$. Take the interval
$[-n\epsilon, n\epsilon]$, and construct
$m$ thresholds at intervals of size $2 n \epsilon / m$; let
the $j$th such threshold be
\begin{equation*}
  t_j \defeq -n \epsilon + \frac{2 n (j-1) \epsilon}{m}
\end{equation*}
Then we ``assign'' observations to each pair of thresholds, so that
threshold $j$ corresponds to observations $I_j \defeq \{\frac{n (j - 1)}{m}
+ 1, \ldots, \frac{n j}{m}\}$, of which there are $n/m$. For each index $i
\in I_j$, we set
\begin{equation*}
  B_i = \begin{cases}
    1 & \mbox{if~} X_i \ge t_j \\
    0 & \mbox{otherwise}.
  \end{cases}
\end{equation*}
Then we simply set $\theta_n\init$ to be the \newtext{minimal} threshold for which
$B_i = 0$ for all observations $X_i$ corresponding to that threshold.

Let us now consider the probability that $\theta_n\init$ is substantially
wrong. For notational simplicity, let $U_i = (1 + X_i - \theta) / 2$, so that
the $U_i$ are uniform on $[0, 1]$.
First, note that we always have $\theta_n\init \ge \theta
- \frac{2 n \epsilon}{m}$, because no observations will be below
the appropriate threshold. Let $j\opt$ be the \newtext{largest} index $j$
for which $\theta \newtext{\ge} t_{j\opt}$, and consider
the index sets $I_{j\opt}$, $I_{j\opt + 1}$, and so on.
The event $\theta\init_n \ge t_{j\opt} + \frac{2 n \epsilon}{m}$ may
occur only if for each of the $n/m$ observations in the set
$I_{j\opt + 1}$, we have $U_i \ge \frac{n \epsilon}{m}$. Thus,
\begin{align*}
  P_\theta\left(\theta_n\init \ge \theta + \frac{2 n \epsilon}{m}
  \right) \le
  P_\theta\left(\theta_n\init \ge t_{j\opt} + \frac{2 n \epsilon}{m}
  \right)
  \le \left(1 - \frac{n \epsilon}{m}\right)^\frac{n}{m}
  \le \exp\left(-\frac{n^2\epsilon}{m^2}\right).
\end{align*}
Setting the number of bins $m = \sqrt{n}$,
the resolution $\epsilon = 2 \log n / n$, we obtain
$P_\theta(\theta_n\init \ge t_{j\opt} + 4 \log n / \sqrt{n})
\le n^{-2}$. Thus we have
\begin{align}
  \label{eqn:quality-of-initial-estimate}
  \sup_{|\theta| \le \log n} P_\theta\left(|\theta_n\init - \theta| \ge
  \frac{8 \log n}{\sqrt{n}}\right) \le \frac{1}{n^2}.
\end{align}

The second stage estimator follows roughly the same strategy, except that
the resolution of the bins is tighter. In particular, let us assume that
$|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$, which happens
eventually by inequality~\eqref{eqn:quality-of-initial-estimate}.  (We will
assume this tacitly for the remainder of the argument.)  Consider the
interval $\Theta_n \defeq \theta_n\init + [-\frac{16 \log n}{\sqrt{n}},
  \frac{16 \log n}{\sqrt{n}}]$ centered at $\theta_n\init$; we know that the
interval includes $[\theta - \frac{8 \log n}{\sqrt{n}}, \theta + \frac{8
    \log n}{\sqrt{n}}]$. Without loss of generality we assume $\theta_n\init
= 0$.  Following precisely the same discretization strategy as that for
$\theta_n\init$, we divide $\Theta_n$ into $m$ equal intervals, with
thresholds $t_j = -\frac{16 \log n}{\sqrt{n}} + \frac{32 (j - 1) \log n}{m
  \sqrt{n}}$; let $\epsilon_n = \frac{32 \log n}{m \sqrt{n}}$ be the width of
these intervals.  Then following exactly the same reasoning as above, we
assign indices $I_j = \{\frac{n(j - 1)}{m} + 1, \ldots, \frac{n j}{m}\}$ and
for $i \in I_j$, set $B_i = 1$ if $X_i \ge t_j$. We define $\theta_n$ to be
the \newtext{minimal} threshold $t_j$ for which $B_i = 0$ for all observations $X_i
\in I_j$. Then following precisely the reasoning above, we have (on the
event that $|\theta_n\init - \theta| \le \frac{8 \log n}{\sqrt{n}}$)
\begin{equation*}
  P_\theta(|\theta_n - \theta|
  \ge 2 \epsilon_n)
  \le (1 - \epsilon_n)^\frac{n}{m}
  \le \exp\left(-\frac{n \epsilon_n}{m}\right)
  = \exp\left(-\frac{32 \sqrt{n} \log n}{m^2}\right).
\end{equation*}
Set $m = 4 n^{1/4}$ to obtain the claimed
result~\eqref{eqn:uniforms-are-easy}.
